The Choose-Your-Own-Adventure Calculus (Pearl/Brave New Idea)

### 2025-03-06 paper.tex
Submitted version

\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
\hideLIPIcs
\newcommand{\ident}[1]{\textsf{#1}}
\newcommand{\kvd}[1]{\textsf{\bfseries #1}}
\newcommand{\ddident}[1]{\guillemotleft\ident{#1}\guillemotright}
\newcommand{\sigmaN}{\sigma_0}
\newcommand{\select}{\textnormal{\ident{choose}}}
\newcommand{\suggest}{\textnormal{\ident{suggest}}}
\newcommand{\choices}{\textnormal{\ident{choices}}}
\newcommand{\op}{\textit{op}}
\newcommand{\vect}[1]{\lbrack #1 \rbrack}
\newcommand{\preview}{\textnormal{\ident{preview}}}
\newcommand{\links}{\textnormal{\ident{links}}}
\DeclareMathOperator{\argmax}{arg\,max}

@@

\begin{abstract}
\noindent
TODO: Some of the most remarkable results in mathematics reveal connections between different
branches of the discipline. The aim of this paper is to point out a modest, but still
remarkable, similarity between a range of different interactive programming systems.

~

\noindent
TODO: Say somewhere that this also formalizes things like mixed-initiative interaction and
programming by demonstration

~

\noindent
TODO: It is mainly about giving us a nice way to talk about lots of things in interactive
programming systems and enable a transfer of ideas between them

@@

Multiple interactive programming systems, ranging from code editors for object-oriented programming
languages to data exploration systems and interactive proof assistants, exhibit a remarkably
similar pattern of interaction. They offer the user, who can be a programmer, a data scientist
or a proof writer, a range of choices that the user can select from in order to complete their
program, script or proof. The user can initiate the interaction iteratively, using it to
create and refine a larger part of their program.

There are subtle differences between different implementations of the general pattern. In some
systems, the resulting source code will contain a trace of the choices made by the user.
For example, when choosing an item from a list of class members, the code will contain the member
name. In some systems, the interaction results in a block of code that can be included in the
source file, but does not include a trace of the interaction. For example, invoking a proof search
or case split in Idris \cite{brady-2015-idris} constructs a well-typed program, but leaves no trace
of the command used to construct it.
%
The nature of the generated options also varies. The list of choices may include all possible
options that are valid at a given location, or it may list only a subset of the valid options.
In some cases, it may also include incorrect options as, for example, in auto-completion
for dynamic languages \cite{frolich-2024-autocomplete}.

The aim of this paper is paper is to formally capture the recurring interaction pattern:

\begin{enumerate}
\setlength{\itemsep}{5pt}
\item We motivate the formalism by reviewing four different systems that implement a variation on the
interaction pattern. These include type providers for data access in F\# \cite{syme-2013-inforich},
type providers for data exploration in The Gamma \cite{petricek-2022-thegamma,petricek-2017-dotdriven},
AI assistants for semi-automated data wrangling \cite{petricek-2023-aias} and tooling for
interactive proof assistants \cite{altenkirch-1994-alf,brady-2015-idris,verter-2024-mixed} (Section~\ref{sec:motivation}).

@@

\item The calculus allows us to make the aforementioned subtle differences precise. We define the
notions of \emph{correctness} and \emph{completeness} for the choose-your-own-adventure calculus.
To distinguish the different ways of embedding the interactions in the edited programs, we
also formally define \emph{internal} and \emph{external} mode of system integration.

\item We show that various programmer assistance tools, such as search and AI-based recommendations
can be built on top of the primitives offered by the calculus, showing how the
choose-your-own-adventure calculus supports of transfer of ideas across different kinds of
interactive programming systems.
\end{enumerate}

@@

@@

The following sections review four different instances of the choose-your-own-adventure
interaction pattern. In all of those, an interactive editor offers the user some kind of~a~com\-pletion list during working with the system.

@@

F\# type providers \cite{syme-2013-inforich} are a mechanism for integrating external data
sources into the F\# type system. A type provider is a compiler extension, loaded and
executed at compile-time and at edit-time. It can run arbitrary code to read the structure of
external data and use it to generate a suitable statically-typed representation of the
data, typically as objects with members. Type providers can, for example, infer the type from a
sample JSON \cite{petricek-2016-fsdata} or read a database schema.

The example in Figure~\ref{fig:worldbank} shows a simple type provider for accessing information
from the World Development Indicators database. The provided \ident{wb} object allows the programmer
to access any indicator of any country in the database by choosing an appropriate \ident{[Country]}
and an \ident{[Indicator]} in a chain of members
\ident{wb}.\ident{Countries}.\ident{[Country]}.\ident{Indicator}.\ident{[Indicator]}.
The result is a time series with values for the given indicator and a country. More generally,
the example can be seen as a special case of a type provider for slicing n-dimensional
data cube \cite{petricek-2022-thegamma} -- we choose a fixed value for two of the three
dimensions (country, indicator, time).

When using the type provider, the user types the first line of code and triggers auto-completion
by typing \ident{wb} followed by the dot. The rest of the code is constructed by choosing an
option from a list and typing another dot.\footnote{This interaction pattern has been
lightheartedly called \emph{dot-driven development} by Phil Trelford \cite{seemann-2021-head}.}

\begin{figure}[t]
\includegraphics[width=0.48\textwidth]{fig/thegamma1.png}~~
\includegraphics[width=0.48\textwidth]{fig/thegamma2.png}
\caption{Constructing a query in The Gamma. We count the number of gold and silver medals for
each athlete and sort the data by the number of gold medals.}
\label{fig:thegamma}
\end{figure}

@@

The type provider shown in Figure~\ref{fig:thegamma} lets the user construct an SQL-like query by
repeatedly choosing operations and their parameters \cite{petricek-2017-dotdriven}. It keeps track
of the schema and uses it to generate all possible valid parameters. When sortin data, it generates
an object with two members for each columns -- one for ascending and one for descending sort.
Similarly, the grouping operation first offers all columns as possible grouping keys and then
lets the user choose from a range of pre-defined aggregations (sum, count, average, concatenate).
The system also evaluates the query on the fly, providing a live preview during editing \cite{petricek-2020-live}.

The interaction pattern is the same as before. After the usertriggers auto-completion, they
repeatedly select an operation and its parameters to construct a query. One notable difference
is that the structure of the generated types is potentially infinite (the user can keep adding
further operations) and so the types are generated lazily.

@@

Figure~\ref{fig:aia} shows an example. It uses the datadiff \cite{sutton-2018-datadiff} AI
assistant, running in a Wrattler notebook \cite{petricek-2018-wrattler}, to merge broadband
quality data published by Ofcom for two subsequent years. The format of the CSV files for the
two years differs. Columns were added, removed, renamed and their order has changed. In the
example, we selected 6 columns from the year 2015 and want to find matching data from 2014.

When the AI assistants runs automatically, it correctly maps the numerical columns, but it
incorrectly maps the \ident{Urban.rural} (2014) column to \ident{Nation} (2015). This happens
because both columns are categorical and have three values with similar distribution. A data
analyst can easily spot the mistake. They click the ``+'' button to add a constraint and choose
\ident{Don't match `Urban.rural' and `Nation'} to specify that the two columns should not be matched.
Datadiff then runs again and finds the correct matching.

\begin{figure}[t]
  \includegraphics[width=0.9\textwidth]{fig/aia.png}
  \caption{Using the datadiff AI assistant to reconcile the structure of the two datasets.
    The user is offered a list of constraints to prevent or force matching between specific columns.}
  \label{fig:aia}
\end{figure}

~

The interaction patter is the same as in the previous two cases. The analyst constructs the
correct data transformation by repeatedly choosing from a list of options, until they obtain
the desired result. However, the way the interaction pattern is implemented differs.
First, in the case of type providers, we are gradually constructing a program by adding operations to
a method chain. Now, the AI assistant synthesizes a data transformation (program) and we are
gradually adding constraints to control the synthesis. Second, in the case of type providers,
the completion list offered all possible members of the object. Now, the list offers
constraints recommended by the AI assistant which may not be complete.

\begin{figure}[t]
  \includegraphics[width=0.6\textwidth]{fig/alf.png}
  \caption{Constructing a proof of the transitivity of the $\leq$ relation in the \textsc{Alf}
    editor. The user is offered a range of variables and constructors in scope at the current
    location. \cite{altenkirch-1994-alf}}
  \label{fig:alf}
\end{figure}

@@

Systems like Idris provide key bindings to invoke the completions, but the functionality could
also be offered through a user interface. An example that illustrates this is the interactive editor
for the \textsc{Alf} theorem prover \cite{magnusson-1994-alf}, which is based on the refinement of
an incomplete proof object \cite{altenkirch-1994-alf}. This is illustrated in Figure~\ref{fig:alf}.
The user is proving the transitivity of the $\leq$ relation for Peano arithmetic natural numbers.
They pattern match on the proof argument $p$ and complete the first branch. For the second branch,
they need to fill a hole $?_{p2}$ (called a wildcard in \textsc{Alf}). They trigger a completion
and a pop-up menu shows the available variables and constructors, including \ident{leq\_trans} that
can be used to complete the proof. After choosing \ident{leq\_trans}, two new holes are generated
for its arguments. Those can be, again, filled interactively, by choosing $p_1$ and $p$ from the
completion.

The interaction pattern is again the same. The user repeatedly triggers a completion and uses
it to refine and complete their proof by filling holes. There are subtle differences too. Unlike
with AI assistants, each completion directly refines the proof that the user is editing. Unlike
with type providers, a completion may generate multiple new holes, rather than just adding to a
chain of operations.

@@

\begin{figure}[t]
  \includegraphics[width=0.55\textwidth]{fig/sandblocks.png}
  \caption{Constructing a program using the Sandblocks structure editor \cite{beckmann-2023-all}.
    The user is typing the start of an expression and context menu offers possible production
    rules according to a grammar.}
  \label{fig:sand}
\end{figure}

\subparagraph{Structure editors.}
Structure editors make it possible to construct programs by manipulating the abstract syntax tree
of a program rather than by working with textual representation. The interaction with a structure
editor can be based on menus \cite{teitelbaum-1981-synth}, key bindings \cite{weber-2017-editors},
tiles~\cite{moon-2022-tylr} and a range of other approaches.

Sandblocks~\cite{beckmann-2023-all} is an example of a recent structure editor that is interesting
for multiple reasons. First, the editor is automatically generated from a grammar. Second, the
editor combines keyboard input with context menus (Figure~\ref{fig:sand}). In the above example,
the user is typing code and needs to fill a placeholder for an expression. After they type the
start of the expression, the editor offers a completion list with possible production rules from
the grammar that are valid in the given context and are compatible with the text typed so far. The
user can continue typing (to further refine the recommendation), but they can also select a rule
from the menu. This completes the expression according to the production rule, leaving further
empty holes in the code.

In Sandblocks, the context menus serve more as hints and users of the editor often
construct programs through typing, but the system illustrates the fact that structure editors
can also be based on the interaction pattern captured in this paper. The editor can offer
all possible production rules based on the given grammar and the user could construct an
entire program by choosing one of the rules (and manually entering primitive terminals such as
strings and identifiers).

@@

We can think of the interaction with the system as navigating through a tree structure, starting
from a root and choosing one of the possible branches in each step.\footnote{Serving as another
evidence for the surprising effectiveness of the concept of a tree \cite{nesetril-2005-strom}.}
In the following definition, the key $\choices$ operation can thus be seen as returning
branches of a given node.

@@

The definition is not a programming language calculus in the usual sense in that it does not
define a concrete syntax with reduction rules. It is an abstract algebraic structure that captures
the structure of a system that supports the choose-your-own-adventure interaction pattern.
The definition is close to that of an AI assistant \cite{petricek-2023-aias}, which is written
using a language specific for the data wrangling domain (such as cleaning scripts or input and
output data) but is structurally similar. It is also worth noting that the definition may describe
not just trees, but also graphs with cycles -- a system can return to a previously visited state.
This is not practically useful, but it does not pose a theoretical problem.

\subparagraph{Expression completion.}
One of the subtle questions about the choose-your-own-adventure pattern raised in the introduction
is whether a trace of the interaction can be reconstructed from the source code of an interactively
constructed program. That is, if we follow a sequence of choices $\iota_1, \ldots, \iota_n$ to
construct a program $e$, is it possible to recover the original sequene of choices from the
program $e$.

@@

\vspace{-0.5em}
\raggedright
\begin{enumerate}
\item $\ident{init}(E[?]) = \sigmaN$ obtains the initial state of a choose-your-own-interaction system,
\item $\sigma_n$ is a system state such that $\forall i\!\in\! 1\ldots n\;.\;(\iota_i\mapsto\sigma_i)\in\choices{(\sigma_{n-1})}$,
  i.e.,~the user\\ makes a series of choices resulting in a final state of the system $\sigma_n$,
\item $E[e]$ where $e=\select(\sigma_n)$, i.e.,~the final program is constructed by replacing the hole
  in the completion context with the expression $e$ generated from $\sigma_n$.
\end{enumerate}
\end{definition}

As we will see when we revisit the earlier examples formally, this way of invoking a choose-your-own-adventure
system is used, for example, in the case of interactive theorem provers like \textsc{Alf} or Idris.
In those systems, the user triggers the completion on a proof (program) containing a hole.
They then fill the hole and, possibly iteratively, further holes in the generated proof. The final
expression is embedded in the source code, but it does not indicate what options, identified by
$\iota_1, \ldots, \iota_n$, were selected in the process.

\subparagraph{Reconstructible completion.}
In many choose-your-own-adventure systems, it is possible to reconstruct a trace of the interaction
through which a program was constructed. This is the case with type providers, where a user chooses
a sequence of object members to be accessed and those members directly appear in the source code.
The same would be the case in a completion system for Coq that would offer tactics to apply,
becuase the resulting proof would contain a record of the selected tactics.

@@

If the integration of a choose-your-own-adventure system with a host programming language
uses reconstructible expression completion, we can recover the choices through which the user completed
an expression $e$ in a completion context $E[e]$, assuming they used the interactive system rather
than entering the code directly. This also means that we can reconstruct the final state $\sigma_n$
of the system by starting from $\ident{init}(E[?])$ and following the choices specified by
$\iota_1, \ldots, \iota_n$.

\newpage
\section{Examples}
\label{sec:examples}

We now revisit the four examples from Section~\ref{sec:motivation} and show how they fit the
above formal model. All four examples rely on some domain-specific logic. We describe what
information the logic provides, but do not model it formally. This has been done elsewhere,
in works describing the individual systems.

To show how the model lets us distinguish subtle details of interactive programming systems,
we start with a model of data exploration system that is inspired by The Gamma, but differs in
one notable way. We then discuss type providers more generally and show how to correctly model
The Gamma. We then revisit the remaining two examples.

@@

In The Gamma, the query is hidden from the data analyst. Behind the scenes, the system generates
objects with members and the identifiers designating individual options are the names of those
members. The operation is encapsulated in the code of the accessor of the member. In the
simplified model in this section we ignore this fact. The model presented here directly
generates code that calls the underlying operations. For example, assume that the user makes
the following choices:
\[
\textnormal{\ddident{group data}\,.\,\ddident{by Athlete}\,.\,\ddident{sum Gold}\,.\,\ddident{count all}\,.\,\ddident{then}}
\]
In The Gamma, the individual identifiers become object members and they are included as
a member chain in the generated code. In the following simplified model, the completion
instead fills the hole with an expression representing the operation:
\[
\ident{group}(\texttt{"Athlete"}, \ident{sum}(\texttt{"Gold"}), \ident{count}())
\]
The two approaches have different human-computer interaction trade-offs. In terms of cognitive
dimensions \cite{green-1989-cogdims,blackwell-2003-cogdims}, the latter has a greater closeness
of mapping, while the former is less cognitively demanding to read for a non-programmer.
As discussed in Section~\ref{sec:properties}, the two implementations of the choose-your-own-adventure
interaction pattern also differ in terms of their formal properties.

>>>>

\documentclass[anonymous, a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
\hideLIPIcs
\newcommand{\ident}[1]{\textsf{#1}}
\newcommand{\kvd}[1]{\textsf{\bfseries #1}}
\newcommand{\ddident}[1]{\guillemotleft\ident{#1}\guillemotright}
\newcommand{\sigmaN}{\sigma_0}
\newcommand{\select}{\textnormal{\ident{choose}}}
\newcommand{\suggest}{\textnormal{\ident{suggest}}}
\newcommand{\choices}{\textnormal{\ident{choices}}}
\newcommand{\op}{\textit{op}}
\newcommand{\vect}[1]{\lbrack #1 \rbrack}
\newcommand{\preview}{\textnormal{\ident{preview}}}
\newcommand{\links}{\textnormal{\ident{links}}}
\DeclareMathOperator{\argmax}{arg\,max}

@@

\begin{abstract}
\noindent
Some of the most remarkable results in mathematics reveal connections between different
branches of the discipline. The aim of this paper is to point out a modest, but still
remarkable, similarity between a range of different interactive programming systems.
%
In many programming systems, the user can interactively construct a program by repeatedly
triggering some kind of completion mechanism and choosing one of the offered options. This
is the case with code editors for object-oriented languages (choosing object members), data
exploration environments (choosing an applicable transformation), but also theorem provers
(choosing an applicable tactic) and structure editors (choosing a grammar~rule).

In this paper, we formally capture the essence of this interaction pattern through a small
formal model called \emph{the choose-your-own-adventure calculus}. We show how a wide range
of different systems fits the model. Looking at the examples through a common perspective
reveals multiple subtle differences. To formally capture those, we characterise
basic properties of choose-your-own-adventure systems, resembling those from other areas
of programming language research, including correctness, completeness and uniqueness.
%
We further show how the choose-your-own-adventure calculus can be used as the basis for
formally studying more advanced interaction patterns including mixed-initiative interaction,
AI-based programming assistants and programming by demonstration.

We strongly believe that interaction with programming systems deserves as much attention
as the underlying programming languages. Our work is one step in this direction. It provides
a way of talking about commonalities and subtle differences between multiple interactive
programming systems and enables a transfer of ideas between interactive systems from very
different domains.

@@

Multiple interactive programming systems, ranging from code editors for object-oriented programming
languages to data exploration systems, interactive proof assistants and structure editors, exhibit a
remarkably similar pattern of interaction. They offer the user, who can be a programmer, a data scientist
or a proof writer, a range of choices that the user can select from in order to complete their
program, script or proof. The user can initiate the interaction iteratively, using it to
create and refine a larger part of their program.

There are subtle differences between different implementations of the general pattern. In some
systems, the resulting source code will contain a trace of the choices made by the user.
For example, when choosing an item from a list of class members, the code will contain the member
name. In some systems, the interaction results in a block of code that can be included in the
source file, but does not include a trace of the interaction. For example, invoking a proof search
or case split in Idris \cite{brady-2015-idris} constructs a well-typed program, but leaves no trace
of the command used to construct it.
%
The nature of the generated options also varies. The list of choices may include all possible
options that are valid at a given location, or it may list only a subset of the valid options.
In some cases, it may even include incorrect options as, for example, in auto-completion
for dynamic languages \cite{frolich-2024-autocomplete}.

The aim of this paper is to formally capture the recurring interaction pattern:

\begin{enumerate}
\setlength{\itemsep}{5pt}
\item We motivate the formalism by reviewing five different systems that implement a variation on the
interaction pattern. These include type providers in F\# \cite{syme-2013-inforich},
type providers for data exploration in The Gamma \cite{petricek-2022-thegamma,petricek-2017-dotdriven},
AI assistants for semi-automated data wrangling \cite{petricek-2023-aias}, tooling for
interactive proof assistants \cite{altenkirch-1994-alf,brady-2015-idris,verter-2024-mixed}
and structure editing \cite{beckmann-2023-all} (Section~\ref{sec:motivation}).

@@

\item The calculus allows us to make the aforementioned subtle differences precise. We define the
notions of \emph{correctness}, \emph{completeness} and \emph{uniqueness} for the choose-your-own-adventure
calculus (Section~\ref{sec:properties}). To understand the way in which the interaction is
integrated in a programming environment, we formally define the notion of \emph{expression
completion} and capture a specfic kind of \emph{reconstructible} completion (Section~\ref{sec:calculus}).

\item We show that multiple programmer assistance tools, such as mixed-initiative interaction,
AI-based assistants and programming by demonstration can be defined on top of the primitives
offered by the calculus (Section~\ref{sec:applications}). This also illustrates how the
choose-your-own-adventure calculus supports transfer of ideas across different kinds of
interactive programming systems.
\end{enumerate}

@@

% ==================================================================================================

@@

We start by reviewing five instances of the interaction pattern. In all of them, an editor
offers the user a completion list to choose from during working with the system.

\begin{figure}[t]
\vspace{-0.5em}
\includegraphics[width=0.48\textwidth]{fig/thegamma1.png}~~
\includegraphics[width=0.48\textwidth]{fig/thegamma2.png}
\caption{Constructing a query in The Gamma. We count the number of gold and silver medals for
each athlete and sort the data by the number of gold medals.}
\vspace{-0.5em}
\label{fig:thegamma}
\end{figure}

% --------------------------------------------------------------------------------------------------

@@

F\# type providers \cite{syme-2013-inforich} are a mechanism for integrating (primarily) external data
sources into the F\# type system. A type provider is a compiler extension, loaded and
executed at compile-time and at edit-time. It can run arbitrary code to read the structure of
external data and use it to generate a suitable statically-typed representation of the
data, typically objects with members. Type providers can, for example, infer the type from a
sample JSON \cite{petricek-2016-fsdata} or read a database schema.

The example in Figure~\ref{fig:worldbank} shows a simple type provider for accessing information
from the World Development Indicators database. The provided \ident{wb} object allows the programmer
to access any indicator of any country in the database by choosing an appropriate \ident{[Country]}
and an \ident{[Indicator]} in a chain of members
\ident{wb}.\ident{Countries}.\ident{[Country]}.\ident{Indicator}.\ident{[Indicator]}.
The result is a time series with values for the given indicator and a country. More generally,
the example can be seen as a special case of a type provider for slicing $n$-dimensional
data cube \cite{petricek-2022-thegamma}. We choose a fixed value for two of the three
dimensions (country, indicator, time) and obtain a series indexed by the remaining dimension.

When using the type provider, the user types the first line of code to initialize the type provider
and triggers auto-completion by typing \ident{wb} followed by the dot. The rest of the code is
constructed by choosing an option from a list and typing another dot (a pattern light-heartedly
called \emph{dot-driven development} by Phil Trelford \cite{seemann-2021-head}).

% --------------------------------------------------------------------------------------------------

@@

The type provider shown in Figure~\ref{fig:thegamma} lets the user construct an SQL-like query by
repeatedly choosing operations and their parameters \cite{petricek-2017-dotdriven}. It keeps track
of the schema and uses it to generate all possible valid parameters. For example, when sorting data, it generates
an object with two members for each columns -- one for ascending and one for descending order.
Similarly, the grouping operation first offers all columns as possible grouping keys and then
lets the user choose from a range of pre-defined aggregations (sum, count, average, concatenate).
The system also evaluates the query on the fly, providing a live preview \cite{petricek-2020-live}.

The interaction pattern is the same as before. After the user triggers auto-completion, they
repeatedly select an operation and its parameters to construct a query. One notable difference
is that the structure of the generated types is potentially infinite (the user can keep adding
further operations) and so the types are generated lazily.

\begin{figure}[t]
  \vspace{-0.5em}
  \includegraphics[width=0.875\textwidth]{fig/aia.png}
  \caption{Using the datadiff AI assistant to reconcile the structure of the two datasets.
    The user is offered a list of constraints to prevent or force matching between specific columns.}
  \label{fig:aia}
\end{figure}

% --------------------------------------------------------------------------------------------------

@@

Figure~\ref{fig:aia} shows an example. It uses the datadiff AI assistant \cite{sutton-2018-datadiff},
running in a Wrattler notebook \cite{petricek-2018-wrattler}, to merge broadband
quality data published by Ofcom for two subsequent years. The format of the CSV files for the
two years differs. Columns were added, removed, renamed and their order has changed. In the
example, we selected 6 columns from the year 2015 and want to find matching data from 2014.

When the AI assistants runs automatically, it correctly maps the numerical columns, but it
incorrectly maps the \ident{Urban.rural} (2014) column to \ident{Nation} (2015). This happens
because both columns are categorical and have three values with similar statistical distribution. A data
analyst can easily spot the mistake. They click the ``+'' button to add a constraint and choose
\ident{Don't match `Urban.rural' and `Nation'} to specify that the two columns should not be matched.
Datadiff then runs again and finds the correct matching.

The interaction patter is the same as in the previous two cases. The analyst constructs the
correct data transformation by repeatedly choosing from a list of options, until they obtain
the desired result. However, the way the interaction pattern is implemented differs.
First, in the case of type providers, we are gradually constructing a program by adding operations to
a method chain. An AI assistant automatically synthesizes a data transformation (program) and we are
gradually adding constraints to control the synthesis. Second, in the case of type providers,
the completion list offered all possible members of the object. Now, the list offers
constraints recommended by the AI assistant which may not be complete.

\begin{figure}[t]
  \vspace{-0.5em}
  \includegraphics[width=0.6\textwidth]{fig/alf.png}
  \caption{Constructing a proof of the transitivity of the $\leq$ relation in the \textsc{Alf}
    editor. The user is offered a range of variables and constructors in scope at the current
    location. \cite{altenkirch-1994-alf}}
  \label{fig:alf}
\end{figure}

% --------------------------------------------------------------------------------------------------

@@

Systems like Idris provide key bindings to invoke the completions, but the functionality could
also be offered through a user interface. An example that illustrates this is the interactive editor
for the \textsc{Alf} theorem prover \cite{magnusson-1994-alf}, which is based on a gradual refinement of
an incomplete proof object \cite{altenkirch-1994-alf}. This is illustrated in Figure~\ref{fig:alf}.
The user is proving the transitivity of the $\leq$ relation for Peano arithmetic natural numbers.
They pattern match on the proof argument $p$ and complete the first branch. For the second branch,
they need to fill a hole $?_{p2}$ (called a wildcard in \textsc{Alf}). They trigger a completion
and a pop-up menu shows the available variables and constructors, including \ident{leq\_trans} that
can be used to complete the proof. After choosing \ident{leq\_trans}, two new holes are generated
for its arguments. Those can be, again, filled interactively, by choosing $p_1$ and $p$ from the
completion.

The interaction pattern is again the same. The user repeatedly triggers a completion and uses
it to refine and complete a proof by filling holes. There are subtle differences too. Unlike
earlier, each completion directly refines the proof that the user is editing;
a completion may also generate multiple new holes, rather than just appending to a
chain of operations.

@@

\begin{figure}[t]
  \vspace{-0.25em}
  \includegraphics[width=0.525\textwidth]{fig/sandblocks.png}
  \caption{Constructing a program using the Sandblocks structure editor \cite{beckmann-2023-all}.
    The user is typing the start of an expression and context menu offers possible production
    rules according to a grammar.}
  \label{fig:sand}
\end{figure}

% --------------------------------------------------------------------------------------------------

\subparagraph{Structure editors.}
Structure editors make it possible to construct programs by manipulating the abstract syntax tree
of a program rather than by working with text. The interaction with a structure
editor can be based on menus \cite{teitelbaum-1981-synth}, key bindings \cite{weber-2017-editors},
tiles~\cite{moon-2022-tylr} and a range of other approaches. A number of those fit
the choose-your-own-adventure pattern.

Sandblocks~\cite{beckmann-2023-all} is an example of a recent structure editor that is interesting
for multiple reasons. First, the editor is automatically generated from a grammar. Second, the
editor combines keyboard input with context menus (Figure~\ref{fig:sand}). In the above example,
the user is typing code and needs to fill a placeholder for an expression. After they type the
start of the expression, the editor offers a completion list with possible production rules from
the grammar that are valid in the given context and are compatible with the text typed so far. The
user can continue typing (to further refine the recommendation), but they can also select a rule
from the menu. This completes the expression according to the production rule, generating further
empty holes that can, in turn, be interactively filled with code.

In Sandblocks, context menus serve mainly as hints and users of the editor often
construct programs through typing, but the system illustrates the fact that structure editors
can also be based on the interaction pattern captured in this paper. The editor can offer
all possible production rules based on the given grammar and the user could construct an
entire program by choosing one of the rules (and manually entering terminals such as
strings and identifiers).

% ==================================================================================================

@@

We can think of the interaction with the system as navigating through a tree structure, starting
from a root and choosing one of the possible branches in each step. %\footnote{Serving as another
%evidence for the surprising effectiveness of the concept of a tree \cite{nesetril-2005-strom}.}
In the following definition, the key $\choices$ operation can thus be seen as returning
branches of a given node.

@@

The definition is not a programming language calculus in the usual sense in that it does not
define a concrete syntax with reduction rules. It is an abstract algebraic description of
the structure of a system that supports the choose-your-own-adventure interaction pattern.
The definition is close to that of an AI assistant \cite{petricek-2023-aias}, which is written
using a language specific for the data wrangling domain (such as cleaning scripts or input and
output data) but is structurally similar. It is also worth noting that the definition may describe
not just trees, but also graphs with cycles. A system where the user can return to a previously
visited state may not be practically useful, but it does not pose a theoretical problem.

% --------------------------------------------------------------------------------------------------

\subparagraph{Expression completion.}
One of the subtle questions about the choose-your-own-adventure pattern raised in the introduction
is whether a trace of the interaction can be reconstructed from the source code of an interactively
constructed program. That is, if we follow a sequence of choices $\iota_1, \ldots, \iota_n$ to
construct a program $e$, is it possible to recover the original sequene of choices through which
it was constructed just from the program $e$.

@@

\vspace{-0.5em}
\raggedright
\begin{enumerate}
\item $\ident{init}(E[?]) = \sigmaN$ obtains the initial state of a choose-your-own-interaction system,
\item $\sigma_n$ is a system state such that $\,\forall i\!\in\! 1\ldots n\;.\;(\iota_i\mapsto\sigma_i)\in\choices{(\sigma_{n-1})}$,
  i.e.,~the user\\ makes a series of choices resulting in a final state of the system $\sigma_n$,
\item $E[e]$ where $e=\select(\sigma_n)$ is the final program constructed by replacing the hole
  in the completion context with the expression $e$ generated from $\sigma_n$.
\end{enumerate}
\end{definition}

As we will see when we revisit the earlier examples formally, this way of invoking a choose-your-own-adventure
system is used, for example, in the case of interactive theorem provers.
In those systems, the user triggers the completion on a proof (program) containing a hole.
They then fill the hole and, possibly iteratively, further holes in the generated proof. The final
expression is embedded in the source code, but it does not indicate what options, identified by
$\iota_1, \ldots, \iota_n$, were selected in the process.

% --------------------------------------------------------------------------------------------------

\subparagraph{Reconstructible completion.}
In many choose-your-own-adventure systems, it is possible to reconstruct a trace of the interaction
through which a program was constructed. This is the case with type providers, where the user chooses
a sequence of object members to be accessed and those members directly appear in the source code.
The same would be the case in a completion system for Coq that would offer tactics to apply,
becuase the resulting proof would consist of a sequence formed by the selected tactics.

@@

If the integration of a choose-your-own-adventure system with a host programming language
uses reconstructible expression completion, we can recover the choices through which the user
constructed the expression $e$ in a completion context $E[e]$. This also means that we can
reconstruct the final state $\sigma_n$ of the system by starting from $\ident{init}(E[?])$
and following the choices specified by $\iota_1, \ldots, \iota_n$.

% ==================================================================================================

\section{Examples}
\label{sec:examples}

We now revisit the five examples from Section~\ref{sec:motivation} and show how they fit the
above formal model. All five examples rely on some domain-specific logic. We describe what
information the logic provides, but refer to full description elsewhere for details.

To show how the model lets us distinguish subtle details of interactive programming systems,
we start with a model of data exploration system that is inspired by The Gamma, but differs in
one notable way. We then discuss type providers more generally and show how to model The Gamma
more precisely. We then revisit the remaining three examples.

% --------------------------------------------------------------------------------------------------

@@

In The Gamma, the query is hidden from the data analyst. Behind the scenes, the system generates
objects with members and the identifiers designating individual options are the names of those
members. The operation is encapsulated in the code of the accessor of the member. In the
simplified model developed in this section we ignore this fact. The model presented here directly
generates code that calls the underlying operations. For example, assume that the user makes
the following choices:
\[
\textnormal{\ddident{group data}\,.\,\ddident{by Athlete}\,.\,\ddident{sum Gold}\,.\,\ddident{count all}\,.\,\ddident{then}}
\]
In The Gamma, the individual identifiers become object members and they are included as
a member chain in the generated code. In the simplified model presented here, the completion
instead fills the hole with an expression representing the operation:
\[
\ident{group}(\texttt{"Athlete"}, \ident{sum}(\texttt{"Gold"}), \ident{count}())
\]
The two approaches have different human-computer interaction trade-offs. In terms of cognitive
dimensions \cite{green-1989-cogdims,blackwell-2003-cogdims}, the latter has a greater closeness
of mapping, while the former is less cognitively demanding to read for a non-programmer.
As we will see in Section~\ref{sec:properties}, the two implementations of the
interaction pattern also differ in terms of their formal properties.

\subparagraph{Formal model.}
The options generated by The Gamma let the user select both the next operation and the parameters
of the previously selected operation. The available operations and parameters are generated
based on a schema $S$ that is transformed by the operations. The state of the system $\sigma$
contains the current schema $S$ and the operations applied so far. In the following, we write
$\op(\boldsymbol{p})$ for an operation with a vector of parameters:
\[
\sigma = S, \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n})}\\
\]
The behaviour of the $\choices$ operation depends on whether the last operation in the sequence
expects further parameters or whether it is fully-specified. In the first case, the recommendation
engine generates possible additional parameter values $p', p'', \ldots$ based on the schema $S$,
the operation $\op_n$ and the already known parameters $\boldsymbol{p_n}$.
The $\choices$ operation then generates options that add the additional parameter. We generate the
identifiers $\iota',\iota'',\ldots$ based on the state and the parameter value, such as
\ddident{by Gold descending}. Note that adding a parameter may also result in a new schema
$S', S'', \ldots$ (which the recommendation engine computes based on the previous schema and the new parameter):
\[
\begin{array}{l}
\choices(S, \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n})}) =\\
\qquad \{\; \iota' \mapsto (S', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n},p')}), \\
\qquad\;\;    \iota'' \mapsto (S'', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n},p'')}), ~\ldots~\}
\end{array}
\]
If the last operation takes no further parameters, the system produces a choice
of possible next operations $\op', \op'', \ldots$. Again, we are also given new schemas $S', S'', \ldots$
and we generate identifiers $\iota',\iota'',\ldots$ based on the operation name. The $\ident{choices}$
operation then returns options that append the additional operation:
\[
\begin{array}{l}
\choices(S, (\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n}))) =\\
\qquad \{\; \iota' \mapsto (S', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n}), \op'()}), \\
\qquad \;\; \iota'' \mapsto (S'', \vect{\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n}), \op''()}), ~\ldots~\}
\end{array}
\]
Finally, the $\select$ operation takes the state $\sigma$ and generates an expression that represents
the data transformation. This is only possible if all parameters are fully-specified. For simplicity,
assume that $k$ is the index of the last fully-specified operation (either $n$ or $n-1$). If the
host language lets us compose functions using $f\circ g$, we can write:
\[
\select(S, (\op_1(\boldsymbol{p_1}), \ldots, \op_n(\boldsymbol{p_n}))) = \op_1(\boldsymbol{p_1}) \circ \ldots \circ \op_k(\boldsymbol{p_k})
\]
The recommendation engine behind The Gamma provides a domain-specific logic for generating
possible operations and their parameters based on the current schema of the data. As the above
definition shows, this underlying engine can be wrapped and exposed through the common
choose-your-own-adventure interface.
